---
title: "23. Case study on survival analysis"
author: "dizhen"
date: "5/1/2020"
output: 
  html_document:
    toc: yes
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# library(Design)
library(mice)
library(foreign)
library(survival)
library(dplyr)
library(Hmisc)
```


Importation of SMART-data-file with correct variables and endpoints
```{r}
# Variables are already truncated to 1th en 99th percentiles
# SMART  <- read.spss('data/SMARTst.sav',use.value.labels=F, to.data.frame=T)

# Original SMARTfile not truncated
# SMARTo  <- read.spss('data/SMARTs.sav',use.value.labels=F, to.data.frame=T)
load("data/SMART.RData")
load("data/SMARTo.RData")
glimpse(SMART)
glimpse(SMARTo)
```

## PAPER STEP 1: Data Inspection

```{r}
# Understand the data
SMART[1,]
dim(SMART)
describe(SMART)

# Generate an extra window for 1 plot
par(mfrow=c(1,1))

# Selecting 18 variables out of 29 in the loaded SMART-file and immediatly retrieve info again.
SMARTm <- SMART[,c("SYSTBP", "DIASTBP", "SYSTH", "DIASTH","CHOL","HDL",
                   "LDL","TRIG", "HOMOC", "GLUT", "CREAT", "IMT", "albumin",
                   "STENOSIS", "DIABETES", "SMOKING", "packyrs", "alcohol")]
describe(SMARTm)

```

### Cluster analysis of patterns of missingness

```{r}
# With the following function you determine the fraction of NA's (missing values)
# Cluster analysis of patterns of missingness in the SMART study (n = 3873)

na.patterns <- naclus(SMARTm)
# Plot the 18 variables you just selected in a graph
plot(na.patterns, ylab="Fraction of NAs in common")
```

We note that the diastolic and systolic blood pressure measurements are always jointly missing.

```{r}
# A different approach to show the NA's is the following:
# Patterns of missing data in the SMART study.

par(mfrow=c(2,2))
naplot(na.patterns, which=c('na per var'))
naplot(na.patterns, which=c('na per obs'))
naplot(na.patterns, which=c('mean na'))
naplot(na.patterns, which=c('na per var vs mean na'))

```

### Multiple imputation

```{r}
# (though multiple imputation is preferable from a theoretical view point, single imputation was considered more practical and sufficient to obtain reasonable predictions.)

(fmla <- as.formula(paste(" ~ ", paste(colnames(SMARTm), collapse=" +"))))

a <- aregImpute(fmla, n.impute = 2,nk=0, data = SMARTm)
# It is difficult to know selection with descrete. You have to force your variables to act linearly(nk=0) or force selected variables to be linear using I(xj).

par(mfrow=c(4,5))
plot(a)

```


```{r}
# Multiple imputation

# SMARTm <- SMART[,c("SYSTBP", "DIASTBP", "SYSTH", "DIASTH","CHOL","HDL",
                   # "LDL","TRIG", "HOMOC", "GLUT", "CREAT", "IMT", "albumin",
                   # "STENOSIS", "DIABETES", "SMOKING", "packyrs", "alcohol")]
imputed <-impute.transcan(a, data=SMART, imputation=1, list.out=TRUE, pr=FALSE, check=FALSE)


SMART[names(imputed)] <- imputed
head(SMART)


na.patterns <- naclus(SMART)
# Plot the 18 variables you just selected in a graph
plot(na.patterns, ylab="Fraction of NAs in common")



```


## PAPER STEP 2: Coding of Predictors.

We used the Cox regression model. This model is appropriate for prediction of an outcome at relatively short-term such as 5 year cumulative incidence of cardiovascular events. For long-term predictions (e.g. 10 year incidences), a parametric model might be preferable such as a Weibull model. A Weibull model provides more stable estimates at the end of the follow-up.


### Extreme Values

Before any modelling started, the distributio of all potential predictors were carefully examined for extreme values. Biologically implausible values were set to missing values, and remaining extreme values were truncated by shifting the values below the 1 centile and above the 99 centile to "truncation points".

```{r}
# Showing the maximum value of IMT in the truncated file.
max(SMART$IMT,na.rm=T)  
# Show how many times this maximum value is exceeded in the original file
sum(SMARTo$IMTO>1.83,na.rm=T)
# Same, but now for the minimum value
min(SMART$IMT,na.rm=T)  
sum(SMARTo$IMTO<.47,na.rm=T)

```



```{r}
# show relation between IMT in both the not-truncated file and the truncated file with the cox proportional hazard function
coxph(Surv(TEVENT,EVENT) ~ IMT, data=SMART)
coxph(Surv(TEVENT,EVENT) ~  IMTO, data=SMARTo)
```


```{r}
library(rms)
library(survminer)
library(riskRegression)


# again an extra screen
par(mfrow=c(1,1),mar=c(5,5,3,1))

# boxdf <- data.frame(Truncated = SMART$IMT, Origianl = SMARTo$IMTO)
# boxplot(boxdf, ylab = "IMT (mm)")

# make an new matrix with the the original and truncated IMT variable. Then name the two colums
IMTdata <- as.matrix(cbind(SMARTo$IMTO, SMART$IMT))
IMTlist <- split(IMTdata, col(IMTdata))
names(IMTlist)  <- Cs(Original, Truncated)
# make a boxplot of this matirx to show the difference of the data when it's truncated 
boxplot(IMTlist, ylab= "IMT (mm)", cex.lab=1.2)  
```

```{r}
# Now the effect on outcome is shown with the cph-function in three ways 
#to show how the variable which fit the model best: 
#1: Not truncated with restricted cubic spline with 5 knots
dd <- datadist(SMARTo)
options(datadist="dd")
IMTOfit <- cph(Surv(TEVENT,EVENT) ~  rcs(IMTO,5), data=SMARTo)
pred1 <- Predict(IMTOfit)
plot(pred1, xlab="IMT (mm)", lty=2, lwd=2, conf.int=F)
# plot(Predict(IMTOfit), xlab="IMT (mm)", lty=2, lwd=2, conf.int=F)
pred1$type <- rep("IMTO",length(Predict(IMTOfit)[,1]))

#2: Truncated with restricted cubic spline with 5 knots 
dd <- datadist(SMART)
options(datadist="dd")
IMTfit <- cph(Surv(TEVENT,EVENT) ~  rcs(IMT,5), data=SMART)
pred2 <- Predict(IMTfit) 
plot(pred2, xlab="", lty=3, lwd=2,conf.int=F)
pred2$type <- rep("IMT",length(Predict(IMTfit)[,1]))

#3: Truncated linear
IMTfitlin <- cph(Surv(TEVENT,EVENT) ~  IMT, data=SMART)
pred3 <- Predict(IMTfitlin)
plot(pred3, xlab="",  lty=1, lwd=2, conf.int=F)
pred3$type <- rep("IMTlin",length(Predict(IMTfitlin)[,1]))


names(pred1) <- c("x","yhat","lower","upper",".predictor.","type")
names(pred2) <- c("x","yhat","lower","upper",".predictor.","type")
names(pred3) <- c("x","yhat","lower","upper",".predictor.","type")
preddf <- rbind(pred1, pred2, pred3)[,c(1:6)]
preddf <- as.data.frame(preddf)
ggplot(preddf, aes(x = x, y = yhat, color = type)) +
  geom_line( lwd=1)+
  labs(x = "IMT (mm)", y = "log Relative Hazard")+ 
  scale_color_discrete(name = "IMT", labels = c("Truncated Spline", "Truncated Linear", "Original Spline")) 
```

### Transforming continuous predictors


We consider several transformations of variables. Based on the Wald $\chi^2$, we know how the model fit or the performance with those covariates. 

Results: A transformation such as (Age-50)^2+ led to an better model with $chi^2 = 130$. For creatinine, a log transformation gave the best fit. A linear coding of systolic blood pressure was reasonable and diastolic blood pressure had no effect when we analyzed the convertional sphygmomanometry blood pressure measurements ("SYSTH" and "DIASTH" variables). 


Start of Age example.
```{r}
# Example transformation of age to get best fit.
# With cph-function the X2 and dof are measured when the variable in transformed.
dd <- datadist(SMART)
options(datadist="dd")
# par(mfrow=c(2,2), mar=c(5,5,3,1))
# par(mfrow=c(1,1))
# age related to outcome: age linear --> X2=97
AGEfit <- cph(Surv(TEVENT,EVENT) ~  AGE, data=SMART)  
plot(Predict(AGEfit), 
     xlab="Age (years)", ylab = "log Relative Hazard",
     lty=1, lwd=2,conf.int=T, 
     ylim=c(-1.5, 1.5), xlim=c(20,80))
df1 <- Predict(AGEfit)
df1 <- df1 %>% mutate(type = rep(1,length(df1[,1])))
anova(AGEfit)

# age related to outcome: age + age squared --> X2= 125
AGEfit2 <- cph(Surv(TEVENT,EVENT) ~  pol(AGE,2), data=SMART)  # age + age squared
df2 <- Predict(AGEfit2)
df2 <- df2 %>% mutate(type = rep(2,length(df2[,1])))
anova(AGEfit2)
plot(Predict(AGEfit2), 
     xlab="Age (years)", ylab = "log Relative Hazard",
     lty=2, lwd=2,conf.int=T
     )

plotdf <- rbind(df1, df2)

plotdf %>%ggplot(aes(x = AGE, y = yhat, color = factor(type))) +
  geom_line(lwd=1)+
  # geom_point() + 
  geom_ribbon(aes(ymin=lower, ymax=upper), linetype=2, alpha=0.1)+
  labs(title = paste("Age, 1 df, ", "chi^2"," ", 97, "; Age+","Age^2",", 2 df, ", "chi^2"," ", 125,sep=""), 
       y = "log Relative Hazard",
       x = "Age (years)")+
  scale_color_discrete(name = "Transformation", labels = c("Age", "Age + Age^2"))
  
```


```{r}
# Age related to outcome: 125truncated under 50 / 55 and then linear --> X2= 119
AGEfit3 <- cph(Surv(TEVENT,EVENT) ~  ifelse(AGE>55, (AGE-55),0), data=SMART)  # age linear
anova(AGEfit3)
df3 <- Predict(AGEfit3)
df3 <- df3 %>% mutate(type = rep(3, length(df3[,1])))
plot(Predict(AGEfit3), 
     xlab="Age (years)",ylab = "log Relative Hazard",
     lty=1, lwd=2,conf.int=T, 
     ylim=c(-1.5, 1.5), xlim=c(20,80))

# Age related to outcome: 125truncated under 50 / 55 and then squared --> X2= 130
AGEfit4 <- cph(Surv(TEVENT,EVENT) ~  ifelse(AGE>50, (AGE-50)^2,0), data=SMART)  # age + age squared
anova(AGEfit4)
df4 <- Predict(AGEfit4)
df4 <-df4 %>% mutate(type = rep(4, length(df4[,1])))
plot(Predict(AGEfit4), 
     xlab="Age (years)", ylab = "log Relative Hazard",
     lty=2, lwd=2,conf.int=T)


plotdf <- rbind(df3, df4)

plotdf %>%ggplot(aes(x = AGE, y = yhat, color = factor(type))) +
  geom_line(lwd=1)+
  # geom_point() + 
  geom_ribbon(aes(ymin=lower, ymax=upper), linetype=2, alpha=0.1)+
  labs(title = paste("Age>55, 1 df, ", "chi^2"," ", 119, "; ","(Age>50)^2",", 2 df, "," chi^2"," ", 130,sep=""), 
       x = "Age (years)",
       y = "log Relative Hazard")+
  scale_color_discrete(name = "Transformation", labels = c("Age>55", "(Age>50)^2"))

```

```{r}
# age related to outcome with Spline and 3 df (4 knots) --> X2=125
AGEfit5 <- cph(Surv(TEVENT,EVENT) ~  rcs(AGE,4), data=SMART)  
anovaAge  <- anova(AGEfit5)
plot(Predict(AGEfit5), xlab="Age (years)", lty=3, lwd=2,conf.int=T, 
     ylim=c(-1.5, 1.5), xlim=c(20,80))
# title(substitute(paste("Spline, 3 df, ", "chi^2"," ", 125,sep="")), cex.main=1)
plotdf <- as.data.frame(Predict(AGEfit5))

plotdf %>% ggplot(aes(x = AGE, y = yhat)) +
  geom_line(lwd=1)+
  # geom_point() +
  geom_ribbon(aes(ymin=lower, ymax=upper), linetype=2, alpha=0.1)+
  labs(title = paste("Spline, 3 df, ", "chi^2"," ", 125,sep=""),
       x = "Age (years)",
       y = "log Relative Hazard")
```

```{r}
# Age related to outcome in quartiles --> X2=93
tapply(SMART$AGE, ifelse(SMART$AGE<50,1,ifelse(SMART$AGE<60,2,ifelse(SMART$AGE<70,3,4))), mean)
SMART$AGEcat  <- as.factor(ifelse(SMART$AGE<50,43,ifelse(SMART$AGE<60,55,ifelse(SMART$AGE<70,65,74))))
# SMART$AGEcat <- as.factor(SMART$AGEcat)
dd <- datadist(SMART)
options(datadist="dd")
AGEfit6 <- cph(Surv(TEVENT,EVENT) ~  AGEcat, data=SMART)

anova (AGEfit6)
plotdf <- as.data.frame(Predict(AGEfit6))
plot(Predict(AGEfit6), 
     xlab="Age (years)", ylab = "log Relative Hazard",
     lty=3, lwd=2,conf.int=T, 
     ylim=c(-1.5, 1.5))
# title(substitute(paste("Categorized, 3 df, ", "chi^2"," ", 93,sep="")), cex.main=1)

plotdf %>% ggplot(aes(x = AGEcat, y = yhat)) +
  geom_pointrange(aes(ymin=lower, ymax=upper)) +
  labs(title = paste("Categorized, 3 df, ", "chi^2"," ", 93,sep=""),
       x = "Age (years)",
       y = "log Relative Hazard")
```

```{r}
# Age related to outcome dichotomization --> X2=72
tapply(SMART$AGE, cut2(SMART$AGE,g=2), mean)
AGEfit7 <- cph(Surv(TEVENT,EVENT) ~  ifelse(SMART$AGE<61,51,69), data=SMART) 
anova(AGEfit7)
##End AGE-illustration
```


```{r}
# Transformations of other continuous vars: RCS with 3 or 4 knots and normal linear is checked
# BMI
fit <- cph(Surv(TEVENT,EVENT) ~  rcs(BMI,4), data=SMART)  
anova(fit)
fit <- cph(Surv(TEVENT,EVENT) ~  rcs(BMI,3), data=SMART)  
anova(fit) 

# Blood pressure: diastolic
fit <- cph(Surv(TEVENT,EVENT) ~  rcs(DIASTH,4), data=SMART)  
anova(fit)
fit <- cph(Surv(TEVENT,EVENT) ~  rcs(DIASTH,3), data=SMART)  
anova(fit) 
fit <- cph(Surv(TEVENT,EVENT) ~  DIASTH, data=SMART)  
anova(fit) 

# Blood pressure: Systolic
fit <- cph(Surv(TEVENT,EVENT) ~  rcs(SYSTH,4), data=SMART)  
anova(fit)
fit <- cph(Surv(TEVENT,EVENT) ~  rcs(SYSTH,3), data=SMART)  
anova(fit) 
fit <- cph(Surv(TEVENT,EVENT) ~  SYSTH, data=SMART)  
anova(fit) 

# Blood pressure: other type of measurement: diastolic
fit <- cph(Surv(TEVENT,EVENT) ~  rcs(DIASTBP,4), data=SMART)  
anova(fit)
fit <- cph(Surv(TEVENT,EVENT) ~  rcs(DIASTBP,3), data=SMART)  
anova(fit) 
# plot(Predict(fit))
fit <- cph(Surv(TEVENT,EVENT) ~  DIASTBP, data=SMART)  
anova(fit) 

# Blood pressure: other type of measurment: systolic
fit <- cph(Surv(TEVENT,EVENT) ~  rcs(SYSTBP,4), data=SMART)  
anova(fit)
fit <- cph(Surv(TEVENT,EVENT) ~  rcs(SYSTBP,3), data=SMART)  
anova(fit) 
fit <- cph(Surv(TEVENT,EVENT) ~  SYSTBP, data=SMART)  
anova(fit) 

# Lipids: Cholesterol  
fit <- cph(Surv(TEVENT,EVENT) ~  rcs(CHOL,4), data=SMART)  
anova(fit)
fit <- cph(Surv(TEVENT,EVENT) ~  rcs(CHOL,3), data=SMART)  
anova(fit) 
fit <- cph(Surv(TEVENT,EVENT) ~  CHOL, data=SMART)  
anova(fit) 

# Lipids: HDL 
fit <- cph(Surv(TEVENT,EVENT) ~  rcs(HDL,4), data=SMART)  
anova(fit)
fit <- cph(Surv(TEVENT,EVENT) ~  rcs(HDL,3), data=SMART)  
anova(fit) 
fit <- cph(Surv(TEVENT,EVENT) ~  HDL, data=SMART)  
anova(fit) 

# Lipids: LDL
fit <- cph(Surv(TEVENT,EVENT) ~  rcs(LDL,4), data=SMART)  
anova(fit)
fit <- cph(Surv(TEVENT,EVENT) ~  rcs(LDL,3), data=SMART)  
anova(fit) 
fit <- cph(Surv(TEVENT,EVENT) ~  LDL, data=SMART)  
anova(fit) 

# Lipids: Triglyceriden 
fit <- cph(Surv(TEVENT,EVENT) ~  rcs(TRIG,4), data=SMART)  
anova(fit)
fit <- cph(Surv(TEVENT,EVENT) ~  rcs(TRIG,3), data=SMART)  
anova(fit) 
fit <- cph(Surv(TEVENT,EVENT) ~  TRIG, data=SMART)  
anova(fit) 

# Only HDL seems to have an effect
# Other: Homocysteine 
fit <- cph(Surv(TEVENT,EVENT) ~  rcs(HOMOC,4), data=SMART)  
anova(fit)
fit <- cph(Surv(TEVENT,EVENT) ~  rcs(HOMOC,3), data=SMART)  
anova(fit) 
fit <- cph(Surv(TEVENT,EVENT) ~  HOMOC, data=SMART)  
anova(fit) 

# Other: glucose 
fit <- cph(Surv(TEVENT,EVENT) ~  rcs(GLUT,4), data=SMART)  
anova(fit)
fit <- cph(Surv(TEVENT,EVENT) ~  rcs(GLUT,3), data=SMART)  
anova(fit) 
fit <- cph(Surv(TEVENT,EVENT) ~  GLUT, data=SMART)  
anova(fit) 

# Other: Diabetes  
fit <- cph(Surv(TEVENT,EVENT) ~  DIABETES, data=SMART)  
anova(fit)

# Other: Diabetes + Glucose  
fit <- cph(Surv(TEVENT,EVENT) ~  DIABETES + GLUT, data=SMART)  
anova(fit) 
#Best is to keep diabetes, no glucose

```

```{r}
#Other: Creatinine: Also visualiised  in plots, because different transformation is done (logistic)
#Plot 4 graphs
# par(mfrow=c(1,1))
fit <- cph(Surv(TEVENT,EVENT) ~  CREAT, data=SMART)  
anova(fit) 
plot(Predict(fit), 
     main = paste("Linear, 1 df, ", "chi^2"," ", 93,sep=""),
     xlab="Creatinin (mmol/l)", ylab = "log Relative Hazard",
     lty=1, lwd=2,conf.int=T, ylim=c(-.5, 2.5))
# title(substitute(paste("Linear, 1 df, ", chi^2," ", 93,sep="")), cex.main=1.2)

fit <- cph(Surv(TEVENT,EVENT) ~  rcs(CREAT,4), data=SMART)  
anova(fit)
plot(Predict(fit), 
     main = paste("Spline, 3 df, ", "chi^2"," ", 116,sep=""),
     xlab="Creatinin (mmol/l)", ylab = "log Relative Hazard",
     lty=3, lwd=2,conf.int=T, ylim=c(-.5, 2.5))
# title(substitute(paste("Spline, 3 df, ", chi^2," ", 116,sep="")), cex.main=1.2)

fit <- cph(Surv(TEVENT,EVENT) ~  rcs(CREAT,3), data=SMART)  
anova(fit) 
plot(Predict(fit), 
     main = paste("Spline, 2 df, ", "chi^2"," ", 99,sep=""),
     xlab="Creatinin (mmol/l)", ylab = "log Relative Hazard", 
     lty=2, lwd=2,conf.int=T, ylim=c(-.5, 2.5))
# title(substitute(paste("Spline, 2 df, ", chi^2," ", 99,sep="")), cex.main=1.2)

fit <- cph(Surv(TEVENT,EVENT) ~  log(CREAT), data=SMART)  
anova(fit) 
plot(Predict(fit), 
     main = paste("Log transformation, 1 df, ", "chi^2"," ", 131,sep=""),
     xlab="Creatinin (mmol/l)", ylab = "log Relative Hazard", 
     lty=1, lwd=2,conf.int=T, ylim=c(-.5, 2.5))
# title(substitute(paste("Log transformation, 1 df, ", chi^2," ", 131,sep="")), cex.main=1.2)
# Hence: log(CREAT) good choice
```

### Combining predictors with similar effects

Combining predictors with similar effects can be an effective way to limit the degrees of freedom of predictors in a model, but may decrease the model performance.

## PAPER STEP 3: Model Specification

We Look at the Hazard Ratio (HRs) and conrtibution to Cox regression model($chi^2$ and df) of predicors in a full model for cardiocascular events in the SMART study.

To select predictors:
  
  1. A backward selection procedure with Akaike's Information Criterium (AIC). AIC applies a p value < 0.157 for selection of predictors with 1 df.
  
  2. The Lasso method which achieves selection of predictors by shrinking some coefficients to zero by setting a constraint on the sum of the absolute standardized coefficients. It is optimal when there are 10 predictors.

```{r}
# imputation model with outcome, all potential predictors, and auxilliary variables
load("data/SMART.RData")
SMARTM  <-aregImpute(~I(TEVENT)+EVENT+SEX+I(AGE)+
                       SYSTBP+DIASTBP+SYSTH+DIASTH+
                       DIABETES+CEREBRAL+CARDIAC+AAA+PERIPH+
                       # I(HISTCARD)+
                       STENOSIS+ I(LENGTH)+I(WEIGHT)+I(BMI)+
                       I(CHOL)+I(HDL)+I(LDL)+I(TRIG)+I(HOMOC)+I(GLUT)+I(CREAT)+I(IMT)+
                       as.factor(albumin) +as.factor(SMOKING) + I(packyrs) + as.factor(alcohol), 
                     n.impute=5, data=SMART) # all predictors

# Check results of imputation
SMARTM
# Use transcan to make 1 singly imputed file.
SMARTc <- SMART
imputed <-impute.transcan(SMARTM, imputation=1, data=SMART, list.out=TRUE, pr=FALSE, check=FALSE)
SMARTc[names(imputed)] <- imputed
SMARTc$HISTCAR2  <- SMARTc$CEREBRAL+SMARTc$CARDIAC+2*SMARTc$AAA+SMARTc$PERIPH  
# SMARTc$HISTCAR3  <- ifelse(SMARTc$HISTCARD==4, 3, SMARTc$HISTCARD)
SMARTc[1:4,]

#Check and investigate your new file, no missings anymore
describe(SMARTc)
```


```{r}
# 3 models: full model, LASSO-model, Backward-step model

#1: Full model combined
dd <- datadist(SMARTc)
options(datadist="dd") 

## factor or as.factor is never supposed to work with rms.
# cox01 <-cph(Surv(TEVENT,EVENT) ~  ifelse(AGE>50, (AGE-50)^2,0) + SEX + 
#               as.factor(SMOKING) + as.factor(alcohol)+BMI + SYSTH +  
#               HDL+DIABETES +HISTCAR2 +HOMOC + log(CREAT)+ as.factor(albumin)+STENOSIS+IMT,
#               data = SMARTc, x=T, y=T, surv=T)

cox01 <-coxph(Surv(TEVENT,EVENT) ~ ifelse(AGE>50, (AGE-50)^2,0) +
                SEX +BMI + SYSTH + factor(alcohol)+ as.factor(SMOKING)+
                HDL+DIABETES +HISTCAR2 +HOMOC + log(CREAT)+
                as.factor(albumin)+STENOSIS+IMT,
              data = SMARTc
            )
anova(cox01)
summary(cox01)
cox01
# slightly different numbers because of new impuation      



# Interactions with sex, 10 df, chi2 14.8
anova(coxph(Surv(TEVENT,EVENT) ~  SEX * (ifelse(AGE>55, (AGE-55),0) + 
                                           BMI + HDL+DIABETES +HISTCAR2 + log(CREAT)+
                                           as.factor(albumin)+STENOSIS+IMT),
            data = SMARTc))

# specific sex*HISTCAR2, 1 df
anova(coxph(Surv(TEVENT,EVENT) ~  ifelse(AGE>55, (AGE-55),0) + 
              BMI + HDL+DIABETES +SEX * HISTCAR2 + log(CREAT)+
              as.factor(albumin)+STENOSIS+IMT,
            data = SMARTc))

# HISTCAR2 * other predictors, 9 df
anova(coxph(Surv(TEVENT,EVENT) ~  SEX + HISTCAR2 * (ifelse(AGE>55, (AGE-55),0) +
                                                      BMI + HDL+DIABETES + log(CREAT)+
                                                      as.factor(albumin)+STENOSIS+IMT),
            data = SMARTc))

# specific sex* +CEREBRAL+CARDIAC+AAA+PERIPH
anova(coxph(Surv(TEVENT,EVENT) ~  ifelse(AGE>55, (AGE-55),0) + 
              BMI + HDL+DIABETES +SEX*(CEREBRAL+CARDIAC+AAA+PERIPH) + log(CREAT)+
              as.factor(albumin)+STENOSIS+IMT,
            data = SMARTc))


# Plot interaction history * sex
par(mfrow=c(1,2))
# levels(SMARTc$SEX) <- Cs(Male, Female)
dd <- datadist(SMARTc)
options(datadist="dd")
modfit <- coxph(Surv(TEVENT,EVENT) ~  ifelse(AGE>55, (AGE-55),0) + 
           BMI + HDL+DIABETES +SEX*HISTCAR2 +log(CREAT)+ as.factor(albumin)+STENOSIS+IMT,data = SMARTc)
modfit

# plot(Predict(coxph(Surv(TEVENT,EVENT) ~  ifelse(AGE>55, (AGE-55),0) + 
#                      BMI + HDL+DIABETES +SEX*HISTCAR2 +log(CREAT)+
#                      as.factor(albumin)+STENOSIS+IMT,data = SMARTc)),
#      HISTCAR2=NA, SEX=NA, xlim=c(1,5.8), 
#      xlab="Sumscore previous symptomatic atherosclerosis")
# title('sex*sumscore')
# text(x=c(5.2,5.2), y=c(0.93,2.56), Cs(Male, Female), adj=0)

# C-statistic of the full model:
rcorr.cens(-cox01$linear.pred,cox01$y) 
```

```{r}
cox01 <-coxph(Surv(TEVENT,EVENT) ~ ifelse(AGE>50, (AGE-50)^2,0) +
                SEX +BMI + SYSTH + factor(alcohol)+ as.factor(SMOKING)+
                HDL+DIABETES +HISTCAR2 +HOMOC + log(CREAT)+ as.factor(albumin)+STENOSIS+IMT,
              data = SMARTc
            )

#2: Stepwise selection
library(MASS)
reg.model <- stepAIC(cox01,direction = "backward")
summary(reg.model)

# works for cph but not coxph
# backwardstepwise <- fastbw(cox01, rule="aic", type="individual")
# backwardstepwise
# cox01.step  <- update(cox01, ~ cox01$x[,backwardstepwise$parms.kept])
# cox01.step

```
```{r}
cox01 <-coxph(Surv(TEVENT,EVENT) ~ ifelse(AGE>50, (AGE-50)^2,0) +
                SEX +BMI + SYSTH + factor(alcohol)+ as.factor(SMOKING)+
                HDL+DIABETES +HISTCAR2 +HOMOC + log(CREAT)+
                as.factor(albumin)+STENOSIS+IMT,
              data = SMARTc
            )

#3: LASSO-method
## glmpath analysis; 'lasso'
library(glmpath)
memory.size(2048)

y <- Surv(SMARTc$TEVENT,SMARTc$EVENT)
x <- model.matrix(y ~ ifelse(AGE>50, (AGE-50)^2,0) +
                SEX +BMI + SYSTH + factor(alcohol)+ as.factor(SMOKING)+
                HDL+DIABETES +HISTCAR2 +HOMOC + log(CREAT)+
                as.factor(albumin)+STENOSIS+IMT,
              data = SMARTc)


## SMART data analysis
SMARTdata <- list(x=x, time=SMARTc$TEVENT, status=SMARTc$EVENT)
summary(SMARTdata)

## VERY LONG computing time ...
# predictie corrector methode voor berekenen van CPH-model met L1 penalty

# SMARTpath <- coxpath(data=SMARTdata) 
# save(SMARTpath, file = "data/SMARTpath.RData")
load("data/SMARTpath.RData")

## Or alternatives
# library("glmnet")
# SMARTpath <- glmnet(x, Surv(TEVENT,EVENT) ~ ifelse(AGE>50, (AGE-50)^2,0) +
#                 SEX +BMI + SYSTH + factor(alcohol)+ as.factor(SMOKING)+
#                 HDL+DIABETES +HISTCAR2 +HOMOC + log(CREAT)+
#                 as.factor(albumin)+STENOSIS+IMT, family = "cox")

par(mar=c(5,4,4,6))
plot.coxpath(SMARTpath, cex.lab=1.2)
plot.coxpath(SMARTpath, cex.lab=1.2,type = "aic")
SMARTpath
SMARTpath$aic
round(SMARTpath$b.predictor[SMARTpath$aic==min(SMARTpath$aic),],4)

```


## PAPER STEP 4: Model Estimation

To illutrate the statistics:

1. Regression coefficient were first estimated as default with Cox regression analysis - by maximizing the log-likelihood of the fit of the model to the data.

2. The coefficients of the nine predictors in the stepwise backward selected model were rather similar to their corresponding coefficients in the full model.

3. In contrast, the lasso model shrunk coefficients of weaker predictors such as BMI, HDL, diabetes and albumin considerably towards zero

To illutrate two figures above:

1. AIC: Lasso path with increasing sum of the absolute standardized coefficients ($|\beta|$). The optimal AIC is obtained with 11 predictors, but differences are small between models with 9-12 predictors. 

2. coefficient path: predictors have effects other than zero with higher $|\beta|$.


```{r}
###CHECK SMARTHPATH AND CHOOSE ROW IN WHICH HOMOCYST IS DELETED FROM THE MODEL###
SMARTpath$b.predictor
# model ?? is without homocyst
Lasso.coefs <- SMARTpath$b.predictor[1,];Lasso.coefs
Lasso.coefs <- Lasso.coefs[Lasso.coefs > 0 | Lasso.coefs < 0]
Lasso.coefs

Lasso.coefs <- SMARTpath$b.predictor[2,];Lasso.coefs
Lasso.coefs <- Lasso.coefs[Lasso.coefs > 0 | Lasso.coefs < 0]
Lasso.coefs

Lasso.coefs <- SMARTpath$b.predictor[3,];Lasso.coefs
Lasso.coefs <- Lasso.coefs[Lasso.coefs > 0 | Lasso.coefs < 0]
Lasso.coefs

Lasso.coefs <- SMARTpath$b.predictor[26,];Lasso.coefs
Lasso.coefs <- Lasso.coefs[Lasso.coefs > 0 | Lasso.coefs < 0]
Lasso.coefs

```

```{r}
cox01 <-coxph(Surv(TEVENT,EVENT) ~ ifelse(AGE>50, (AGE-50)^2,0) +
                SEX +BMI + SYSTH + factor(alcohol)+ as.factor(SMOKING)+
                HDL+DIABETES +HISTCAR2 +HOMOC + log(CREAT)+
                as.factor(albumin)+STENOSIS+IMT,
              data = SMARTc
            )

y <- Surv(SMARTc$TEVENT,SMARTc$EVENT)
x <- model.matrix(y ~ ifelse(AGE>50, (AGE-50)^2,0) +
                SEX +BMI + SYSTH + factor(alcohol)+ as.factor(SMOKING)+
                HDL+DIABETES +HISTCAR2 +HOMOC + log(CREAT)+
                as.factor(albumin)+STENOSIS+IMT,
              data = SMARTc)


# Alternsative: Goeman's library
library(penalized)
memory.size(2048)

## new SMART data analysis
opt <- optL1(y, penalized = x, fold = 10, model='cox', maxlambda1=20, standardize=T)
coefficients(opt$fullfit)

cox01.lasso <- penalized(y ~ x, model='cox', standardize=T, lambda1=18)
coef(cox01.lasso)

cox01.lasso.path <- penalized(y ~ x, model='cox', standardize=T, lambda1=1, steps=10)

plotpath(cox01.lasso.path, standardize=T)
title('Standardized coefficients')
```

## PAPER STEP 5: Model performance

1. Model validation: Stability

* We use bootstrap resampling procedure to study the stability of our stepwise selected model, and to quantify the optimism of our modelling strategy. 

2. Model validation: Optimism

* We use c statistic.


```{r}
## PAPER STEP 5: Model performance ##

# cox01 <-coxph(Surv(TEVENT,EVENT) ~ ifelse(AGE>50, (AGE-50)^2,0) +
#                 SEX +BMI + SYSTH + factor(alcohol)+ as.factor(SMOKING)+
#                 HDL+DIABETES +HISTCAR2 +HOMOC + log(CREAT)+
#                 as.factor(albumin)+STENOSIS+IMT,
#               data = SMARTc
#             )



#COX01DEF:
cox01def <-coxph(Surv(TEVENT,EVENT) ~  ifelse(AGE>50, (AGE-50)^2,0) +
                   BMI + HDL + DIABETES +HISTCAR2 + log(CREAT)+
                   as.factor(albumin)+STENOSIS+IMT,
                 data = SMARTc,x = T, y = T) # x=T, y=T,surv=T
dd <- datadist(SMARTc)
options(datadist="dd")
anova(cox01def)
summary(cox01def)

#C-statistic and KM-curves of cox01def
#C-statistic:
rcorr.cens(-cox01def$linear.pred,cox01def$y) 
#KM-curves:
# validate(cox01def, B=20, dxy=T) # validate can only apply to rms::cph

cox01def$linear.pred
summary(cox01def$linear.predictor)
SMARTc$lp4	<- cut2(as.numeric(cox01def$linear.predictor), g=4)
dd <- datadist(SMARTc)
options(datadist="dd")

table(SMARTc$lp4)
levels(SMARTc$lp4) <- as.character(1:4)

KMrisk4	<- cph(Surv(TEVENT,EVENT) ~ lp4, data=SMARTc, x = T, y = T, surv = T)
KMrisk4      

par(las=1)
survplot(KMrisk4, n.risk=T, xlab="Follow-up (days)", cex.n.risk=1,ylab="Fraction free of cardiovascular event", time.inc=365)

```

```{r eval=FALSE}
#Bootstrap:
val.step  <- validate(cox01, bw=T, rule="aic", type="individual")
vcox01def <- validate.cph(cox01def, method='boot', B=200, bw=T,type="individual",dxy=T)
coef.sh 	<- cox01def$coefficients*vcox01def["Slope","index.corrected"]
lp.s		<-cox01def$x%*%coef.sh
rcorr.cens (-lp.s,Surv(SMARTc$TEVENT,SMARTc$EVENT))

# SHRINKAGE FACTORS 
Lasso.coefs <- SMARTpath$b.predictor[SMARTpath$aic==min(SMARTpath$aic),]
Lasso.coefs <- Lasso.coefs[SMARTpath$b.predictor[SMARTpath$aic==min(SMARTpath$aic),] > 0 | 
                             SMARTpath$b.predictor[SMARTpath$aic==min(SMARTpath$aic),] < 0]
Lasso.coefs
cox01def$coef
round(Lasso.coefs/cox01def$coef,3)
mean(Lasso.coefs/cox01def$coef)        # 0.80; stronger shrinkage for weaker effects
# v Houwelingen shrinkage factor, with 17 df --> 0.94
(cox01def$stats[3] -  cox01$stats[4]) / cox01def$stats[3]

```

```{r eval = False}
# Nomogram with lasso coefs as in the original paper ( EERST cox01def draaien, daarvoor moet worden geimputeerd).
lassocoefs <- c(AGE=0.0012,BMI=-0.001,HDL=-0.16,DIABETES=0.11,HISTCAR2=0.33,CREAT=0.71,'ALBUMIN=2'=0.13,'ALBUMIN=3'=0.20,STENOSIS=0.16,IMT=0.50)
cox01def$coefficients <- lassocoefs
surv	<- Survival(cox01def)
surv3	<- function(x) surv(3*365.25,lp=x)
surv5	<- function(x) surv(5*365.25,lp=x)
nomogram(cox01def, fun=list(surv3, surv5), lp=F, AGE=c(15,seq(50,85,5)), CREAT=c(60,80,100,150,200,400),
         funlabel=c('3-year survival', '5-year survival'), maxscale=10, lp.at=c(0,1,2,2.4), fun.at=c(.93,.9,.85,.8,.7,.6,.5))


# nomogram ( if you run this before you run the above, you get the nomogram based on the coefficients of the cph (cox01def) fit) 
surv	<- Survival(cox01def)
surv1	<- function(x) surv(1*365.25,lp=x)
surv3	<- function(x) surv(3*365.25,lp=x)
surv5	<- function(x) surv(5*365.25,lp=x)

par(mfrow=c(1,1))
nomogram(cox01def, fun=list(surv3,surv5), lp=F, AGE=c(15,seq(50,85,5)), CREAT=c(60,80,100,150,200,400), funlabel=c('3-year survival', '5-year survival'), maxscale=10,lp.at=c(0,1,2,2.4), fun.at=c(.93,.9,.85,.8,.7,.6,.5))

```

```{r eval=FALSE}
## EXTRA
# how I computed the absolute risks per patient

#make baseline database
baseline <- data.frame(AGE=0,BMI=0,HDL=0,DIABETES=0,HISTCAR2=0,CREAT=1,ALBUMIN=1,STENOSIS=0,IMT=0)
summary(survfit(cox01def, newdata=baseline))
plot(survfit(cox01def, newdata=baseline))

#  make new variables in which tranformed variables are transformed.                   
SMARTc_new<-SMARTc
#New columns in SMARTc, columns 35-38:
SMARTc_new$ALBUMIN2<-as.numeric(SMARTc$ALBUMIN==2)
SMARTc_new$ALBUMIN3<-as.numeric(SMARTc$ALBUMIN==3)
SMARTc_new$logCREAT<-log(SMARTc$CREAT)
SMARTc_new$newAGE<-ifelse(SMARTc$AGE>50, (SMARTc$AGE-50)^2,0)

##IMPORTANT: NUMBERS ARE THE NUMBERS OF THE COLOMNS IN ORDER OF THE COEFFICIENTS### THIS CAN CHANGE PER DATASET
SMARTc_new[1,]
cox01def$coefficients
# you can give  the coefficients every value you like it to have : see above ...$coefficients<- .....
linpred<-as.matrix(SMARTc_NEW[,c(38,17,19,5,32,37,35,36,10,25)])%*%cox01def$coefficients # depends wheterh you have overwritten the coeffcients if lasso-beta's are used or not  

#compute risks

base1yr<-min(survfit(cox01def, newdata=baseline)$surv[survfit(cox01def, newdata=baseline)$time<=365.25])
base3yr<-min(survfit(cox01def, newdata=baseline)$surv[survfit(cox01def, newdata=baseline)$time<=1095.75])
base5yr<-min(survfit(cox01def, newdata=baseline)$surv[survfit(cox01def, newdata=baseline)$time<=1826.25])

SMARTc$oneyrsurv <- base1yr^(exp(linpred))
SMARTc$threeyrsurv <- base3yr^(exp(linpred))
SMARTc$fiveyrsurv <- base5yr^(exp(linpred))
SMARTc$linpred <- linpred



## SAVE IMAGE ##
save(list = ls(all=TRUE), file = "#choose name#.RData") 
```




